---
title: "Chattanooga Bike Data Exploratory analysis"
author: "Sandra Kilpatrick and Kate Ergle"
date: "September 22, 2015"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: default
---

```{r, echo = FALSE, warning=FALSE, message=FALSE}
rm(list=ls())
#setwd("/Users/sandrakilpatrick/bike_data")
setwd("C:/Users/Marc/Desktop/datascience_chat/github/bikeData/")
options(stringsAsFactors = FALSE)
require(ggplot2)
require("seqinr") # for the col2alpha()
require(gridExtra) # for the grid.arrange()
require(scales) # for date_breaks() function
```


```{r echo = FALSE}
load("data/df.Rdata")
```

# Number of trips

## Per year

```{r, echo = FALSE}
table(df$yearStart)
```

## Per month

```{r, echo = FALSE, fig.width=9, fig.height=4}
# define a new column with for the start Date with a format YY-MM-01
df$StartMonthYear <- format(df$StartDate, "%Y-%m-%01")

# Group the data frame by the newly created column 'StartMonthYear'
StartMonthYearDf <- as.data.frame(table(df$StartMonthYear))
names(StartMonthYearDf) <- c("StartMonth", "Total")
# Need to have the date in a correct format so it would work when plotting
StartMonthYearDf$StartMonth <- as.POSIXct(StartMonthYearDf$StartMonth)

g <- ggplot(data = StartMonthYearDf, aes(x = StartMonth, y = Total)) +
    geom_line() +
    scale_x_datetime(labels = date_format("%Y-%m"), breaks = date_breaks("months")) +
    theme(axis.text.x = element_text(angle = 45)) +
    xlab("Month of the start date of the trip") +
    ylab("Total number of trips") +
    ggtitle("Number of trips per month")
g
```



# Trip duration

## Boxplots per year

```{r, echo = FALSE, fig.width=10, fig.height=3, warning=FALSE}
g1 <- ggplot(df, aes(factor(yearStart), tripDurationHours)) + 
  stat_boxplot(geom ='errorbar') +
  geom_boxplot(outlier.colour = col2alpha("black",0.15)) +
  xlab("Years") +
  ylab("Hours") +
  ggtitle("Trip durations in hours per year")

ylim1 = boxplot.stats(df$tripDurationHours)$stats[c(1, 5)]

g2 <- ggplot(df, aes(factor(yearStart), tripDurationHours)) +
  stat_boxplot(geom ='errorbar') +
  geom_boxplot(outlier.size=NA, varwidth=T) +
  coord_cartesian(ylim = ylim1 * 1.05) +
  xlab("Years") +
  ylab("Hours") +
  ggtitle("Trip durations in hours per year without outliers") 

##Warning: Removed 2057 rows containing missing values (geom_point).
## Warning: Removed 1884 rows containing missing values (geom_point).
## Warning: Removed 693 rows containing missing values (geom_point).

grid.arrange(g1, g2, ncol=2)
```

## Summary per year

```{r, echo = FALSE, fig.width=4, fig.height=2}
tapply(df$tripDurationHours, df$yearStart, summary)
```

## Entries per member type

```{r, echo = FALSE}
table(df$MemberType)
```

**Customer** - a bike user who has bought one or three day pass.   
**Subscriber** - a bike user who has an annual pass.   
**Dependent** - No information.   

## Boxplots per member type  

```{r, echo = FALSE, fig.width=10, fig.height=3, warning=FALSE}

g1 <- ggplot(df, aes(factor(MemberType), tripDurationHours)) + 
  stat_boxplot(geom ='errorbar') +
  geom_boxplot(outlier.colour = col2alpha("black",0.15)) +
  xlab("Member type") +
  ylab("Hours") +
  ggtitle("Trip durations in hours per member type")

ylim1 = boxplot.stats(df$tripDurationHours)$stats[c(1, 5)]

g2 <- ggplot(df, aes(factor(MemberType), tripDurationHours)) +
  stat_boxplot(geom ='errorbar') +
  geom_boxplot(outlier.size=NA, varwidth=T) +
  coord_cartesian(ylim = ylim1 * 1.05) +
  xlab("Member type") +
  ylab("Hours") +
  ggtitle("Trip durations in hours per member type without outliers")

##Warning: Removed 2057 rows containing missing values (geom_point).
## Warning: Removed 1884 rows containing missing values (geom_point).
## Warning: Removed 693 rows containing missing values (geom_point).

grid.arrange(g1, g2, ncol=2)
```

# Ideas

1. The top stations
2. Map
3. Per month and per year the usage of the bikes
4. Check if the shooting had an influence on this
5. Customers vs Subscribers (Trip duration)
